---
title: "Reproducible Research - Course Project 1"
author: "Antonello Finucci"

output: md_document
---


## Loading and preprocessing the data

The following code download the data directly from the link given on the course website and
save it in the R working directory. 

The data are stored in the DF "Activity_Data".

Please note that in this DF the "NAs" have been removed.
```{r, echo=TRUE}
          
            
            zipUrl <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
            zipFile = 'activity.zip'
            if (!file.exists(zipFile)){
                download.file(zipUrl, zipFile, method = 'curl')
            }
            filePath = 'Course_Project_01' 
            if (!file.exists(filePath)){
                unzip(zipFile)
            }
            
            
            # load activity data
            Activity_Data <- read.csv("activity.csv")
            
            
            Activity_Data <-na.omit(Activity_Data)
            library(lubridate)
            Activity_Data$date <- ymd(Activity_Data$date)
```


This is an overview of the first 10 rows of the DF and its structure:
```{r echo=TRUE}
            head(Activity_Data,10)
            str(Activity_Data)

```

## What is mean total number of steps taken per day?

The next chunk of code use the package "dplyr" to store in a new DF called "Activity_Data_day" the total steps per day. This DF is used afterward to create the Histogram at point 2.

```{r, echo=TRUE}

          library(dplyr)

          Activity_Data_day <-group_by(Activity_Data,date)%>%
                              arrange(desc(date))%>%
                              summarize(steps = sum(steps))
```

This is an extract of the first 10 rows of the DF "Activity_Data_day"

```{r, echo=TRUE}

          head(Activity_Data_day, 10)
          

```


```{r scatterplot, echo=TRUE}

          
          hist(Activity_Data_day$steps, main = NULL , xlab = "steps")

```

The following code compute the mean and the median of the steps taken per day.

```{r, echo=TRUE}
            
            steps_mean <- mean(Activity_Data_day$steps)
            steps_mean <- format(steps_mean, scientific = FALSE, digits = 0)
            steps_median <- median(Activity_Data_day$steps)
            

```

The mean of steps taken each day is `r steps_mean`

The median of steps taken each day is `r steps_median`

## What is the average daily activity pattern?

```{r, fig.width=10, echo=TRUE}
            Activity_mean_day <-group_by(Activity_Data, interval)%>%
                                summarize(steps = mean(steps))

            library(ggplot2)

            g <- ggplot(Activity_mean_day, aes(interval,steps)) + geom_line()+
               
               labs(y = "Average number of steps")
            print(g)
            
      
```

The next code is used to assess which of the 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.

```{r, echo=TRUE}
            Activity_mean_interval <- data.frame()
            temp<-c()
            j=1
            for(i in seq(0,2355, 5)){
              
              temp<-  filter(Activity_Data, interval== i)
              Activity_mean_interval[j,1]<- mean(temp$steps)
              Activity_mean_interval[j,2] <- i
              j<-j+1
              
            }
            
            int_filtered <- filter(Activity_mean_interval, Activity_mean_interval$V1 == max(Activity_mean_interval$V1, na.rm=TRUE))
            
      

```

The 5 minutes interval with the max number of steps is the one with the indentifier `r int_filtered[1,2]` and the associated average number of steps are `r format(int_filtered[1,1], digits = 0)`.

## Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset

```{r , echo=TRUE}

          Activity_Data_imputed <- read.csv("activity.csv")
          Activity_Data_imputed$date <- ymd(Activity_Data_imputed$date)
          
```


The dataset contains a total of `r sum(is.na(Activity_Data_imputed$steps))` NAs values.
The total number of observations is `r length(Activity_Data_imputed$steps)` that means `r format(sum(is.na(Activity_Data_imputed$steps)) / length(Activity_Data_imputed$steps) * 100, digits = 0)`% are missing data. Because this is not a small percentage the effects of missing data on the results needs to be assessed.

### 2. Devise a strategy for filling in all of the missing values in the dataset.

The following strategy is used for imputing missing data.

If there is a missing data "NA" the mean of the interval to which it belong is used instead.


### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in

In the following code chunck a for loop and an if are used to replace all the "NAs" present in the dataset.

```{r, echo=TRUE}
            for(i in 1:length(Activity_Data_imputed$steps)){
                if(is.na(Activity_Data_imputed[i,1])){
                    Activity_Data_imputed[i,1]<- Activity_mean_interval[Activity_Data_imputed[i,3] == 
                                                                  Activity_mean_interval$V2, 1]
                }
              
            }
```


Hereafter are shown the begin and the end of the new dataset "Activity_Data" where no "NAs" are present!

```{r, echo=TRUE}
            head(Activity_Data_imputed, 10)
            tail(Activity_Data_imputed, 10)
            str(Activity_Data_imputed)
    

```

As a final check the total number of "NAs" in the dataset  is computed using the function `sum(is.na(Activity_Data_imputed$steps)) is equal to `r sum(is.na(Activity_Data_imputed$steps))`.


### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.



```{r, echo=TRUE}

          Activity_Data_day_i <-group_by(Activity_Data_imputed,date)%>%
                              arrange(desc(date))%>%
                              summarize(steps = sum(steps))

          hist(Activity_Data_day_i$steps,col = "red", main = NULL , xlab = "steps")
          
          hist(Activity_Data_day$steps,col = "green", main = NULL , xlab = "steps", add = TRUE)
          
          legend("topright", c("Imputed", "Non-imputed"), col=c("red", "green"), lwd=10)
          
          
          
          
          
          steps_mean <- mean(Activity_Data_day_i$steps)
          steps_mean <- format(steps_mean, scientific = FALSE, digits = 0)
          steps_median <- median(Activity_Data_day_i$steps)
          steps_median <- format(steps_median, scientific = FALSE, digits = 0)
          


```

The new mean of steps taken each day is `r steps_mean`

The new median of steps taken each day is `r steps_median`.

The strategy used to input the missing values does not produce any change on the mean and a minimum one on the median.


What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r, echo=TRUE}

                  Activity_Data_day_2 <-group_by(Activity_Data_imputed,date)%>%
                                                arrange(desc(date))%>%
                                                summarize(steps = sum(steps))

                  head(Activity_Data_day_2, 10)
                  head(Activity_Data_day, 10)
                  
                  summary(Activity_Data_day_2$steps)
                  summary(Activity_Data_day$steps)
                  
          

```

Adding missing values using the above strategy has caused only changes in the 1st and 3rd Quantiles for the total daily number of steps.

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE, cache=TRUE}

          Activity_Data$daytype <- 0

          for(i in 1:length(Activity_Data$date)){

            if(weekdays(Activity_Data[i,2]) == "Saturday"){
           
              Activity_Data[i,4] <- "weekend"
             
            }
            else if(weekdays(Activity_Data[i,2]) == "Sunday"){
           
              Activity_Data[i,4] <- "weekend"
             
            }
            else if(weekdays(Activity_Data[i,2]) != "Saturday"){
              Activity_Data[i,4] <- "weekday"
            }
            else if(weekdays(Activity_Data[i,2]) != "Sunday"){
              Activity_Data[i,4] <- "weekday"
            }
          }

          
          
          Activity_Data$daytype <- as.factor(Activity_Data$daytype)
          
          Activity_Data <-group_by(Activity_Data, interval, daytype)%>%
                              summarize(steps = mean(steps))
          
         
          p <- ggplot(Activity_Data, aes(interval, steps)) +
            geom_line() + facet_wrap(~daytype, nrow = 2, ncol = 1)
          
          print(p)
          
          
          
          

```

